---
title: "R Tutorial - Baby Name Data"
author: "Jon Wheeler"
date: "2022-11-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting to Know R and RStudio

### The Console

R is an interpreted language. Code does not have to be compiled in order to 
run - instead we use an interpreter (vanilla R or the console in RStudio) 
to execute code on the fly.

We can demonstrate this by using the console as a calculator.

```{r intro_console}
12 + 12
12 * 12
12 ** 2
144 - 12
144 / 12
```
The console will also echo strings or simple numeric values. 

```{r console_echo}
12
'twelve'
'Greetings from Earth!'
```
Note that strings need to be enclosed in quotes. Notice what happens when we
try to execute the following:

```{r str_error, eval=FALSE}
twelve
```
The term *object* in R is synonymous with *variable*. The
error message is telling us there is no object or variable named 'twelve' in
the current environment.

#### Exercise: Please Excuse My Dear Aunt Sally

Something to keep in mind as you code calculations is that mathematical
operations in R follow the standard order of operation (PEMDAS).

What will be the result of the following command? Test your answer by running
the command in the console.

How can we change the command to get a result of 42?

```{r exercise_1}
10 - 3 * 6
```

### Objects (Variables)

We assign values to objects in R using the *<-* operator. The keyboard shortcut
for this is **ALT -**. 

```{r assignment}
exponent = 2
gross <- 12 ** exponent
gross_squared <- gross ** exponent
```


