---
title: "R Tutorial - Baby Name Data"
author: "Jon Wheeler"
date: "2022-11-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting to Know R and RStudio

Let's set up a workspace. There are a few different ways of doing this, and
different operating systems may behave differently in terms of navigating
the file system, application permissions, etc.

**Option 1**
1. Download the entire tutorial code repository from GitHub as a zipfile:
<https://github.com/unmrds/cc-R-RStudio/archive/refs/heads/master.zip>
2. Extract the contents of the zipfile to your preferred working directory -
Documents, Desktop, etc.
3. 

### The Console

R is an interpreted language. Code does not have to be compiled in order to 
run - instead we use an interpreter (vanilla R or the console in RStudio) 
to execute code on the fly.

We can demonstrate this by using the console as a calculator.

```{r intro_console}
12 + 12
12 * 12
12 ** 2
144 - 12
144 / 12
```
The console will also echo strings or simple numeric values. 

```{r console_echo}
12
'twelve'
'Greetings from Earth!'
```
Note that strings need to be enclosed in quotes. Notice what happens when we
try to execute the following:

```{r str_error, eval=FALSE}
twelve
```
The term *object* in R is synonymous with *variable*. The
error message is telling us there is no object or variable named 'twelve' in
the current environment.

#### Exercise: Please Excuse My Dear Aunt Sally

Something to keep in mind as you code calculations is that mathematical
operations in R follow the standard order of operation (PEMDAS).

What will be the result of the following command? Test your answer by running
the command in the console.

How can we change the command to get a result of 42?

```{r exercise_1}
10 - 3 * 6
```

### Objects (Variables)

We assign values to objects in R using the *<-* operator. The keyboard shortcut
for this is **ALT -**. 

```{r assignment}
exponent = 2
gross <- 12 ** exponent
gross_squared <- gross ** exponent
```

Objects can also have string values.

```{r str_assignment}
pet <- 'gecko'
```

We can check the data type of an object using the ```class()``` function. We
will look more closely at data types later.

```{r types}
class(exponent)
class(pet)
```

### The Environment Tab

When we previously executed mathematical operations, the output was printed
to the console. Notice that when we assign values to objects, the value of the
object is not printed to the console but is instead displayed in RStudio's
**Environment** tab.

As we develop workflows, it can be helpful to monitor changes to object values.
This is useful especially when we update the values of objects in place.

```{r update_inplace}
age <- 18
age <-  age + 1 # Happy birthday!
age <- age - 4 # This might not be valid.
```

The **Environment** tab provides information about different types of objects
and data structures.

### The Script Pane

So far we have been working in the console, which is great for testing
commands, installing packages, getting help, and other things you may only
want to try out or which don't belong in a script.

Often however, we will do most of our work by running scripts. That's one of
the advantages of using R - scripts allow us to build scalable, reproducible
workflows.

For the remainder of this tutorial, we are mainly going to use an R script 
rather than the console.





